\vspace{-1cm} Model simplicity breeds contempt: using simple models to answer basic questions on species' distributions

### 2021-02-25 manuscript/manuscript.tex
Habemus introduction first draft

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PARAGRAPH 1
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

One of the central goals of ecology is to understand the ways species are distributed across space and time (ref). As a result, ecologists have developed multiple distribution models over the last two decades, to try to untangle the factors that play a role in defining such distributions (ref). These models estimate species' realized niches using several covariates, including environmental variables (ref), species ecological traits' (ref) and phylogenetic relations (ref). For example, ... More recently, some of the focus have shifted towards approaches that estimate and account for biotic factors, such as competitive or facilitative relationships between species. The idea is that by understanding how such biotic and abiotic factors shape species' distributions, we can gain a mechanistic understanding on how ecological communities are established and change over time. However, while some of this approaches can increase the predictive performance of distribution models (ref), the nature of some of the estimates have been theoretically questioned (ref). This was parsimoniously illustrated by \citealt{bla bla}, who ...

@@

Unfortunately, while some of this approaches can increase the predictive performance of distribution models (ref), the nature of some of the estimates have been theoretically questioned (ref). For example...

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PARAGRAPH 2
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Increasing efforts have been devoted to improving the ability of statistical models to predict the presence/absence of species across ranges (ref). Accurately predicting how species are distributed across ranges is crucial for understanding the impacts and effect of global climate change. With that said, accounting for the mechanisms can come at the cost of modelling noise... However, much less attention is paid to how species' distributions compare to each other.

@@

Increasing efforts have been devoted to improving the ability of statistical models to predict the presence/absence of species across ranges. However, much less attention is paid to how species' distributions compare to each other.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PARAGRAPH 3
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% From Ana Norberg: Each statistical modeling method is based on different assumptions that can be viewed as hypotheses about how ecological communities are structured

While the value of gaining a mechanistic understanding of species' distributions is unquestioned, a lot can be learned from shedding light on basic properties of species' realized niches.

Accounting for expert knowledge on species' environmental preferences to understand general distribution patterns.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PARAGRAPH 4
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
Here, we...Bayesian framework... This allow us to account for... as well as to tackle long-standing hypothesis regarding basic aspects of species distributions.

 For non-parametric models, these often do not account for species uncertainties. One can benefit a lot from a bayesian framework. First, it allows us to compare the parameter estimates ... It also help us incorporate prior information.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% PARAGRAPH 5
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
In this work, we first....

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%   Stuff to realocate   %%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

There is no general agreement the shape of species distributions. While many ecological textbooks (Begon et al., 1990, Giller, 1984, Krebs, 1994) assume this to be unimodal and symmetric, some have warned that empirical distributions can take many different forms \citep{austin2002spatial}. There is not an easy way to untangle the true shape of species' distributions, as this shape is likely to showcase idiosyncrasies at the species level and across systems. The aim of this work, it is not to answer these questions nor to provide a general approach that accommodates such idiosyncrasies. Instead, we want to use a model that is solely constrained by the empirical information that we truly have regarding a particular system, relaxing as much as possible the structural constrains of the statistical framework. Then, we want to use this model to answer basic aspects regarding the way systems of many species are distributed along an environmental gradient.

To decide among modelling approaches, we first need to agree on what we know about the system. We know that species occupy a geographic range; therefore, we know that their distributions have finite variance. Indeed, observations on species' geographic variation and optimal climatic conditions have been long documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' distributions. One could point out that we also know that many other factors might influence species' presence/absence---e.g. the influence of biotic interactions among species. However, we do not necessarily have an intuition of how exactly these factors will influence the shape of species' distributions. As a result, if all we truly knew about a species' distribution was that they have finite variance, the most conservative assumption and the safest bet---i.e. the one with the largest entropy---is that such distribution is a Gaussian.

%There is a growing literature on species' distribution models. For the most part, however, these models have often focussed on model performance and understanding the predictive power of certain covariates. This models often take a response variable, such as presence/absence or abundance data, and fit it with a set of environmental variables and species' ecological or phylogenetic relationships. There... The shape of distributions on systems of many species is rarely studied. That is, simple ...
%model comparison... 
%creating maps of relative suitabilities...
%extrapolate to new data...

%It will be argued in this paper that neglect of ecological knowledge is a limiting factor in the application of statistical modelling in ecology and conservation planning

%The field has shifted towards understanding the effects of biotic factors on such distributions. And while the theory is advancing quickly, the...

%This is a generalised additive model.
% Results are presented for both simulated and empirical data.
% But this approach understates the posterior uncertainty in inference. 
%\section*{Introduction}

%%%%%%%%%%%%%%%%%%%%%%%
% Paragraph on Rapoport's rule
%%%%%%%%%%%%%%%%%%%%%%%
% McCain
%Understanding the factors that determine species range size dis- tributions is becoming increasingly urgent as more climate change assessments document the heightened risk for small- ranged species (e.g. Channell & Lomolino, 2000; Sekercioglu et al., 2008; La Sorte & Jetz, 2010; McCain & Colwell, 2011).
%Rapoport’s rule is the positive relationship of species range sizes with increasing latitude (Stevens, 1989), elevation (Stevens, 1992) or water depth (Stevens, 1996). The latitudinal Rapoport’s rule (LRR) is the most examined in the literature, including testing empirical datasets (e.g. Rohde et al., 1993; Roy et al., 1994; Blackburn & Gaston, 1996; Lyons & Willig, 1997; Ruggiero & Lawton, 1998; Reed, 2003; Arita et al., 2005; Hausdorf, 2006), simulation modelling of the expected patterns (Colwell & Hurtt, 1994; Taylor & Gaines, 1999; Case & Taper, 2000; Arita, 2005; Stauffer & Rohde, 2006; Šizling et al., 2009), and theoretical and empirical evidence for the proposed mechanisms (e.g. Rohde, 1992; Kerr, 1999; Taylor & Gaines, 1999; Gaston & Chown, 1999a; Addo-Bediako et al., 2000; Parmesan et al., 2005). The various reviews of the LRR to date suggest that the overall support is weak (e.g. Rohde, 1996; Gaston et al., 1998; Rohde, 1999; Gaston & Chown, 1999b; Ribas & Schoereder, 2006), prin- cipally due to the high degree of variability in the fit to predic- tions (e.g. support: Blackburn & Gaston, 1996; Lyons & Willig, 1997; Price et al., 1997; Arita et al., 2005; no support: Rohde et al., 1993; Roy et al., 1994; Ruggiero & Lawton, 1998; Reed, 2003).
%The elevational Rapoport’s rule (ERR) has received less examination in the literature than the LRR, but there have also been empirical tests of its predictions (e.g. Patterson et al., 1996; Pleguezuelos & Villafranca, 1997; Price et al., 1997; Rahbek, 1997; Fleishman et al., 1998; Ruggiero & Lawton, 1998; Nathan & Werner, 1999; Sanders, 2002; Fu et al., 2004; Chatzaki et al., 2005; Almeida-Neto et al., 2006; Bhattarai & Vetaas, 2006; Hausdorf, 2006) and underlying theory (e.g. Rohde, 1996; Fleishman et al., 1998; Gaston & Chown, 1999a; Hausdorf, 2006; Ribas & Schoereder, 2006). In the ERR, like the LRR, there is a high degree of variability in support from supportive (e.g. Pat- terson et al., 1996; Pleguezuelos & Villafranca, 1997; Price et al., 1997; Rahbek, 1997; Fleishman et al., 1998; Sanders, 2002; Chatzaki et al., 2005; Almeida-Neto et al., 2006; Hausdorf, 2006; Ribas & Schoereder, 2006) to little or no support (e.g. Patterson et al., 1996; Price et al., 1997; Rahbek, 1997; Ruggiero & Lawton, 1998; Nathan & Werner, 1999; Fu et al., 2004; Bhattarai & Vetaas, 2006; Hausdorf, 2006; Ribas & Schoereder, 2006). Similarly, the processes proposed to underlie the ERR (Stevens, 1992) also show a variable amount of theoretical and empirical support (Patterson et al., 1996; Rohde, 1996; Rahbek, 1997; Fleishman et al., 1998; Gaston & Chown, 1999a; Almeida-Neto et al., 2006; Hausdorf, 2006; Colwell, 2011).
Scarce data and little to no attempt to account for uncertainty in the predictions. Similar to rapoport's rule, we can also ask other questions regarding general geographical patterns of species distributions.

@@

%\section*{Discussion}
\clearpage
\bibliography{references}

>>>>

%Consider testing tails of distribution? I would use the Cauchy or generalized normal distribution for heavy tail distribution without skewness; and I would use an exponentially modified Gaussian distribution with heavy tails for a skew distribution with heavy tails.

One of the central goals of ecology is to understand the ways species are distributed across space and time (ref). Over the last two decades, ecologists have developed multiple distribution models to try to untangle the factors that play a role in defining such distributions \citep{guisanPredictiveHabitatDistribution2000, more}. These models estimate species' realized niches using several covariates, including environmental variables \citep{Guisan}, species ecological traits' \citep{pollockRoleFunctionalTraits2012} and phylogenetic relations \citep{ivesGeneralizedLinearMixed2011}. More recently, some of the focus have shifted towards approaches that estimate and account for biotic factors, such as competitive or facilitative relationships between species \citep{ovaskainenHowMakeMore2017}. The idea is that by untangling the ways in which such biotic and abiotic factors shape species' distributions, we can gain a mechanistic understanding on how ecological communities are established and change over time. However, while these factors can increase the predictive performance of some of the models \citep{norbergComprehensiveEvaluationPredictive2019}, the interpretation of the corresponding parameter estimates has been often questioned \citep{gotelliEmpiricalBayesApproach2010, harrisInferringSpeciesInteractions2016, thurmanTestingLinkSpecies2019}. This was best illustrated by \citet{blanchetCooccurrenceNotEvidence2020}, who used basic statistical arguments to highlight the artefactual nature of the link between co-occurrence and species' ecological interactions drawn by some distribution models.
% Other references for biotic interaction=bad: dormannBioticInteractionsSpecies2018

The value of gaining a mechanistic understanding of species' distributions is unquestionable (ref), with several studies highlighting the importance of factors such as biotic interactions and species' dispersal ability in setting their range limits \citep{wiszRoleBioticInteractions2013, pollockUnderstandingCooccurrenceModelling2014, neuschulzBioticInteractionsSeed2018}. That said, a lot can be learned from taking a phenomenological approach, focussing instead on the description of basic properties of species' realized niches. For example, the study of species' range sizes along environmental gradients can reveal general biodiversity patterns that are crucial from a conservation and management perspective \citep{stevensElevationalGradientAltitudinal1992}. Differences in species' responses to the environment could shed light on how climatic processes and historical contingencies have shaped their distributions \citep{rohdeLatitudinalGradientsSpecies1992, more in Rapoport}. Other properties, such as the skewness of species' distributions, can also reveal general underlying processes regarding species' physiological tolerance to different environmental conditions \citep{kaufmanDiversityNewWorld1995}. More generally, understanding the shape of species' realized niches and the extend to which these vary across species is a crucial issue in ecology and biogeography (ref); however, we do not have an effective way to parsimoniously compare the realized niches of many species. Indeed, there is no general agreement on the shape of species' distributions (ref).

Many ecological textbooks (ref) assume the shape of species distributions to be unimodal and symmetric, but some have warned that empirical distributions can take many different forms \citep{austinModelsAnalysisSpecies1987, austin2002spatial}. In practice, distribution frameworks often use logistic regressions with a linear relationship between covariates (but see XX and YY). This is useful because it simplifies the optimization process, but it comes with several statistical shortcomings. First and foremost, such response curve and the linear relationship between covariates often comes with a set of implicit mathematical constrains that might not be biologically justified. From a purely statistical perspective, if all that we are willing to assume is that species occupy finite geographic ranges---i.e. their probability distributions have finite variance---the most conservative statistical approach is to model these as a Gaussian distributions \citep{frankCommonPatternsNature2009}. This is rarely the starting point in most statistical frameworks that study general biodiversity patterns (but see ref), choosing to use instead Gaussian-logit response curves (refs). Other factors might then condition species distributions to showcase heavy-tails or a skewed shapes, revealing interesting ecological processes shaping biodiversity patterns \citep{austinNonlinearSpeciesResponse1976, minchinEvaluationRelativeRobustness1987}. Second, the aforementioned structural constrains also limit our ability to include any prior information to our parameter estimates. Observations on species' geographic variation and optimal climatic conditions have long been documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' realized niches (e.g. \citealt{landoltFloraIndicativaOkologische2010}). That said, this information is rarely accounted for in most modelling approaches, mainly because there is not a straightforward way to feed this information into the parameters of a linear model (\citealt{scherrerEcologicalIndicatorValues2019}; but see \citealt{terbraakWeightedAveragingLogistic1986}). Finally, and perhaps most importantly, a direct biological interpretation of parameter estimates in linear models becomes increasingly difficult as one moves from unimodal and symmetric distributions \citep{jamilGeneralizedLinearMixed2013, terbraakWeightedAveragingLogistic1986} to skewed distributions \citep{huismanHierarchicalSetModels1993}, making the tests of hypothesis on global biodiversity patterns particularly challenging. For example, \citet{huismanHierarchicalSetModels1993} proposed several non-linear models to characterize several features of species' response curves; however, species' environmental indicator values, range size or distribution skewness are difficult to capture following these model structures.

% From Ana Norberg: Each statistical modeling method is based on different assumptions that can be viewed as hypotheses about how ecological communities are structured
% While many ecological textbooks (Begon et al., 1990, Giller, 1984, Krebs, 1994)
%Niche theory predicts that species occurrence and abundance show non-linear, unimodal relationships with respect to environmental gradients (Austin, 1987; Palmer & Dixon, 1990; Whittaker, 1967)
%\citet{} showed that bell curves are not universal response curves in vegetation data, with distributions often presenting skewed responses, and some modelling techniques try to account for this asymmetry in species distributions.

The field of ecology has quickly moved towards mechanistic and process-based approaches to understand species' distributions \citep{wartonManyVariablesJoint2015}. This has resulted in a plethora of models accounting for several biotic and abiotic factors into the predictions of species co-occurrence. Here, we instead rethink traditional modelling approaches and develop a conceptually simple---and yet statistical and computationally complex---statistical framework to revisit some classic hypothesis in ecology and biogeoraphy. In particular, we develop a Bayesian hierarchical model that accounts for all prior information that we have regarding the distribution of alpine plant species along an elevation gradient in the Swiss Alps, including expert knowledge on species environmental indicator values, range sizes, and plant physiology. We start by considering species' response curves as Gaussian distributed, and then we adapt our model to allow for skewed and long-tailed distributions. Using this statistical framework, we are able to compare the basic properties of the realized niches of multiple species, testing for the existence of general biogeographical patterns. First, we test for the Rapopor's rule, which predicts a positive relationship between range size and elevation \citep{stevensElevationalGradientAltitudinal1992}. While this pattern has been largely studied for multiple systems and across gradients \citep{mccainElevationalRapoportRule2013}; contrasting evidence suggests this rule not to be pervasive across species \citep{ribasRapoportEffectWidespread2006, bhattaraiCanRapoportRule2006, mccainElevationalRapoportRule2013}. Our results not only allow us to properly test the existence of this geographical pattern, but they also showcase variation in how different types of species, such as native or neophytes, might respond to an environmental gradient. Second, we study whether or not species' distributions show steeper declines towards stressful conditions, testing the so-called abiotic stress limitation hypothesis (ref). \citet{normandImportanceAbioticStress2009} tested this for vegetation data using \citep{huismanHierarchicalSetModels1993}'s statistical models for several independent species, finding no clear support for such a hypothesis. Our results are able to shed light on this geographical pattern as well as to highlight the degree to which different species will showcase different levels of decline towards stressful conditions. Specifically, we are able to link plant physiological traits to the skewness of their distributions. Overall, we use models that is solely constrained by the empirical information that we truly have regarding a particular system, relaxing as much as possible the structural constrains of the statistical framework. Using these model, we are able uncover the shape of empirical plant distributions and answer fundamental questions regarding the way systems of many species are distributed along environmental gradients.

%%%%%%%%%%%%%%%%%%%%%%%FINAL

% To decide among modelling approaches, we first need to agree on what we know about the system. We know that species occupy a geographic range; therefore, we know that their distributions have finite variance. Indeed, observations on species' geographic variation and optimal climatic conditions have been long documented, with extensive databases compiled by botanists and field ecologists documenting basic knowledge on species' distributions. One could point out that we also know that many other factors might influence species' presence/absence---e.g. the influence of the aforementioned biotic interactions among species. However, we do not necessarily have an intuition of how exactly these factors will influence the shape of species' distributions. Therefore, if all we truly knew about a species' distribution was that they have finite variance, the most conservative assumption and the safest bet---i.e. the one with the largest entropy---is that such distribution is a Gaussian.

%The strength of climate as a range limiting factor will becrucial for the extent to which species will retract from theirsouthern range margins and expand northwards under global warming. occurrence of a species along an environmentalgradient may primarily be limited by its physiological tolerancetowards the abiotically more stressful conditions, while otherfactors, such as biotic interactions, may assume a greater roletowards less stressful conditions (Connell, 1961; Austin, 1990;Brown  et al. , 1996); hereafter referred to as the asymmetricabiotic stress limitation (AASL) hypothesis.

%abiotic stress is generally thought to increasetowards high latitudes and altitudes and to decrease in importancetowards the equatorial lowlands, where biotic interactions havebeen argued to increase in importance (Dobzhansky, 1950;Kaufman, 1995). Accordingly, a geographical prediction of theAASL hypothesis is that abiotic stress may be most importantin upper-latitudinal and upper-altitudinal range boundaries(MacArthur, 1972; Brown  et al. , 1996), but this has received littlestudy. Loehle (1998) suggested that the northern and southernrange limits of North American tree species are determined bycold tolerance and competitive ability, respectively. Similarly,abiotic stress appears to determine the upper-altitudinal rangelimit for some mountainous plant species, while biotic interactionsmay determine the lower limit (Guisan  et al ., 1998; Bruelheide &Scheidel, 1999; Vetaas, 2002). A second aspect of the AASL hypothesis proposes that theshape of species response curves vary along an environmentalgradient: species that occur under the physiologically moststressful conditions show skewed responses with a steep declinetowards the stressful conditions, while species that occur underless stressful conditions show symmetric responses (Austin,1990; also cf. Austin  et al ., 1994; Austin & Gaywood, 1994;Rydgren  et al. , 2003). In line with this, Kaufman (1995) suggeststhat the degree to which a species’ poleward range boundary islimited by abiotic stress increases the further north the species isdistributed. Analogously, we expect increasing stress limitationwith increasing elevation

% Guisan found no evidence of that. 
% No support for Kaufman’s (1995) geographical prediction ofthe AASL hypothesis,

 %There is not an easy way to untangle the true shape of species' distributions, as this shape is likely to showcase idiosyncrasies at the species level and across systems. The aim of this work, it is not to answer these questions nor to provide a general approach that accommodates such idiosyncrasies. Instead, we want to use a model that is solely constrained by the empirical information that we truly have regarding a particular system, relaxing as much as possible the structural constrains of the statistical framework. Then, we want to use this model to answer basic aspects regarding the way systems of many species are distributed along an environmental gradient.

%%%%%%%%%%%%%%% KEY PAPERS AND IDEAS REGARDING HYPOTHESES #################
% EXPLAIN WHY KNOWING THIS BASIC INFORMATION IS IMPORTANT 
% WHY DO YOU CARE
% ABUNDANCE CENTER HYPOTHESIS IDEA
% HOW STEEP THE EDGES ARE RAPIDLY EXPANDING
% species’ thermal tolerance breadths 
% https://eco.confex.com/eco/2020/meetingapp.cgi/Paper/86915
% https://advances.sciencemag.org/content/5/11/eaaz0414
% https://onlinelibrary.wiley.com/doi/full/10.1111/j.1466-8238.2009.00451.x
% https://royalsocietypublishing.org/doi/10.1098/rspb.2010.1295

%%%%%%%%%%%%%%   Species thermal tolerance breadths   %%%%%%%%%%%%%%%%%%%%%
% Species' thermal tolerance increases with latitude (elevation). The asymmetricabiotic stress limitation (AASL) hypothesis
% - mostly macroecological patterns. What about along elevational gradient?
% - Taxonomic scope often restricted. What about plants?
% - upper and lower thermal tolerance data are not always sampled from the same species. Bayes solve the problem?

@@

@@

@@

@@

%\section*{Discussion}
\clearpage
\bibliography{references2}
